\sysfull

### 2017-08-09 paper/abstract.tex

Enabling battery-free devices is a mandatory step towards an environment-friendly Internet of Things (IoT). However, removing the batteries requires IoT to operate on an harvested, ergo intermittent, power supply. This make sustaining long computation very challenging. Correspondingly, there are two approaches to enable long-running computations on intermittenly-powered IoT: (i) checkpointing, where the volatile state of a program is frequently saved to the non-volatile memory, and (ii) task-based approach, where a programmer splits the code into small idempotent modules. Results so far suggest that task-based approach performs better than checkpointing. However, the use of tasks requires that the energy to execute any task must not exceed the maximum limit of the energy buffer. This makes the code unefficient (when executed on a large energy buffer than indended) and potentially completely disables the program (in the opposite case). Therefore, we present \sys: a task-based execution runtime, enabling code portability by means of a new concept of task coalescing. If a code is written for a very small energy buffer, \sys is able to dynamically coalesce these tasks and constructs a bigger (virtual) task to improve the performance. \todo{Re-write the abstract once the paper is ready}{Brandon}

>>>>

Enabling battery-free devices is a mandatory step towards an environment-friendly Internet of Things (IoT). However, removing the batteries requires IoT to operate on a harvested, ergo intermittent, power supply. This makes sustaining long computations challenging. Basically, there are two approaches: (i) checkpointing, where the volatile state of a program is frequently saved to non-volatile memory, and (ii) task-based approach, where a programmer splits the code into small idempotent sections. Results so far suggest that a task-based approach performs better than checkpointing. However, the use of tasks requires that the energy to execute any task must not exceed the capacity of the energy buffer. This makes the code inefficient (when executed on a larger energy buffer than intended) and potentially completely disables the program (in the opposite case). Therefore, we present \sys: a task-based execution runtime, enabling code portability by means of a new concept of task coalescing and automatic compiler-supported task generation. If a code is written for a very small energy buffer, \sys is able to dynamically coalesce these tasks and constructs a bigger (virtual) task to improve the performance. \todo{Re-write the abstract once the paper is ready}{Brandon}

### 2017-08-09 paper/background.tex

Energy-harvesting devices execute software according to the {\em intermittent execution model}~\cite{dino,chain,alpaca,ratchet}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle. Software running on the device executes {\em intermittently} because energy is only available between when the capacitor reaches its threshold and when it is depleted. An intermittent execution is composed of operating periods interspersed with power failures. The frequency of power failures in an intermittent execution depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power that is orders of magnitude less than a device's operating power, making recharging negligible during operation.

Software has different behavior when executed intermittently than when executed with continuous power. Each power failure clears volatile memory (registers, stack, and global variables). Non-volatile memory (e.g., FRAM) persists across power failures. On a power failure, control flows back to a prior point in the execution. By default, after a power failure control flows to the beginning of {\tt main()}. Early work on intermittent computing preserve intermittent progress by periodically capturing and restoring to a non-volatile checkpoint of the volatile execution context, i.e., registers and stack~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using cyclic redundancy check (CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$. Even with the checkpoint at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of a WAR dependence between the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an execution of the code in which power fails during the second loop iteration. The problem arises because line 6 of the execution writes $\texttt{Rmnd}$ in the second iteration, but power fails before capturing the next checkpoint. After restarting from the checkpoint, the update to $\texttt{data}$ at line 10 reads the value of $\texttt{Rmnd}$ that was updated before the power failure, producing an incorrect result.

@@

There are several run time strategies to ensuring task idempotence. One approach, utilized by Alpaca~\cite{alpaca}, is to identify non-volatile data involved in WAR dependences in a task (like DINO~\cite{dino} and Ratchet~\cite{ratchet} did for checkpoints), execute the task using buffered copies of those data, and commit the buffered copies before completing the task. Another strategy is to statically create multiple versions of non-volatile data shared by tasks and ensure by construction that no task reads and writes the same non-volatile memory location~\cite{chain}. Regardless of the idempotence strategy, the key to task-based execution is that fixed, statically defined tasks execute as though atomic, completing after zero or more attempts.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Platform name & Storage capacitor size \\
		\hline \hline
		Moo~\cite{moo} & 0.1\,F \\
		WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
		NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
		NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
		WISP~\cite{wisp5} & 47\,$\mu$F \\
		{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
		{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
		\hline
	\end{tabular} 
	\caption{Comparison of {\em default} energy storage sizes for various battery-less platforms. Observe a huge variation in storage capacity. We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.}
	\label{table:capacitor}
\end{table}

\noindent \textbf{Fixed-size Tasks are Inflexible and Inefficient:} A programmer must write each task with statically defined boundaries, which makes statically sized tasks inflexible and non-portable. The programmer must ensure statically that a task's energy requirements match the device's energy storage capacity. If a programmer writes a program that consumes more energy than the device can buffer, the task will never be able to complete using the buffered energy. The consequence of such tasks is a non-terminating program that repeatedly attempts to re-execute the task of excess energy cost. If a programmer writes a program that consumes less energy than the device can buffer, the system may operate inefficiently. When a task and its successor both complete, the two tasks were interleaved by a task boundary that preserves progress and state (i.e., checkpoint or commit).  If the two tasks were not also interleaved by a power failure, then the computation done to preserve state was {\em unnecessary}, only incurring an overhead on the tasks' executions. Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity that might be \emph{different} for each transiently-powered device (see Table~\ref{table:capacitor}). Such tasks also make it extremely difficult to {\em port} code from  one device to another. A task that is excessively costly on a device with a small energy buffer may become a short, high-overhead task on a device with a larger energy buffer and vice versa. The key problem is that the programmer writes tasks statically for an energy buffer of a fixed size. \sys's {\em task coalescing} mechanism, which we describe in Section~\ref{sec:task_coalescing}, is motivated by the observation that systems with statically defined tasks may be inflexible and inefficient.

@@

\sys is designed for the demands of existing and future energy-harvesting platforms based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. We assume a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform~\cite{wisp} is equipped with a mixture of SRAM and FRAM. Our implementation leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. We do not require a particular non-volatile memory technology, nor do we require architectural additions commodity processors.~\cite{su_date_2017,ratchet,quickrecall,nvp}. \sys supports I/O behavior similar to recent intermittent execution models~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe, asynchronous I/O.

>>>>

Energy-harvesting devices execute software according to the {\em intermittent execution model}~\cite{dino,chain,alpaca,ratchet}.
%
%Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on the device executes {\em intermittently} because energy is only available between when the capacitor reaches its threshold and when it is depleted. An intermittent execution is composed of operating periods interspersed with power failures. The frequency of power failures in an intermittent execution depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power that is orders of magnitude less than a device's operating power, making recharging negligible during operation.

Software has different behavior when executed intermittently than when executed with continuous power. Each power failure clears volatile memory (registers, stack, and global variables). Non-volatile memory (e.g., FRAM) persists across power failures. On a power failure, control flows back to a prior point in the execution. By default, after a power failure control flows to the beginning of {\tt main()}. Early work on intermittent computing preserved intermittent progress by periodically capturing and restoring to a non-volatile checkpoint of the volatile execution context, i.e., registers and stack~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$. Even with the checkpoint at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of a WAR dependence between the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an execution of the code in which power fails during the second loop iteration. The problem arises because line 6 of the execution writes $\texttt{Rmnd}$ in the second iteration, but power fails before capturing the next checkpoint. After restarting from the checkpoint, the update to $\texttt{data}$ at line 10 reads the value of $\texttt{Rmnd}$ that was updated before the power failure, producing an incorrect result.

@@

There are several run time strategies to ensuring task idempotence. One approach, utilized by Alpaca~\cite{alpaca}, is to identify non-volatile data involved in WAR dependences in a task (like DINO~\cite{dino} and Ratchet~\cite{ratchet} did for checkpoints), execute the task using buffered copies of those data, and commit the buffered copies before completing the task. Another strategy is to statically create multiple versions of non-volatile data shared by tasks and ensure by construction that no task reads and writes the same non-volatile memory location~\cite{chain}. Regardless of the idempotence strategy, the key to task-based execution is that fixed, statically-defined tasks execute as though atomic, completing after zero or more attempts.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Platform name & Storage capacitor size \\
		\hline \hline
		%Moo~\cite{moo} & 0.1\,F \\
		WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
		%NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
		NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
		WISP~\cite{wisp5} & 47\,$\mu$F \\
		{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
		{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
		\hline
	\end{tabular} 
	\caption{Comparison of {\em default} energy storage sizes for example battery-less platforms. Observe a huge variation in storage capacity.}
	%We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.
	\label{table:capacitor}
\end{table}

\noindent \textbf{Fixed-size Tasks are Inflexible and Inefficient:} A programmer must write each task with statically defined boundaries, which makes statically-sized tasks inflexible and non-portable. The programmer must ensure statically that a task's energy requirements match the device's energy storage capacity. If a programmer writes a task that consumes more energy than the device can buffer, the task will never be able to complete using the buffered energy. The consequence of such tasks is a non-progressing program that repeatedly attempts to re-execute the task of excess energy cost. If a programmer writes a program that consumes less energy than the device can buffer, the system may operate inefficiently. When a task and its successor both complete, the two tasks were interrupted by a task boundary that preserves progress and state (i.e., checkpoint or commit).  If the two tasks were not also interleaved by a power failure, then the computation done to preserve state was {\em unnecessary}, only incurring an overhead on the tasks' executions. Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity that might be \emph{different} for each transiently-powered device (see Table~\ref{table:capacitor}). Such tasks also make it extremely difficult to {\em port} code from  one device to another. A task that is excessively costly on a device with a small energy buffer may become a short, high-overhead task on a device with a larger energy buffer and vice versa. The key problem is that the programmer writes tasks statically for an energy buffer of a fixed size. \sys's {\em task coalescing} mechanism, which we will describe in Section~\ref{sec:task_coalescing}, is motivated by the observation that systems with statically-defined tasks may be inflexible and inefficient.

@@

\sys is designed for the demands of existing and future energy-harvesting platforms based around general purpose, commodity computing components~\cite{wisp,msp430datasheet}. We assume a device with a memory system that has fast, byte-addressable volatile and non-volatile memory; in particular, our target platform~\cite{wisp} is equipped with a mixture of SRAM and FRAM. Our implementation leverages hardware support for fast, bulk-copying between memories via DMA~\cite{msp430datasheet}. We do not require a particular non-volatile memory technology, nor do we require architectural additions commodity processors~\cite{su_date_2017,ratchet,quickrecall,nvp}. \sys supports I/O behavior similar to recent intermittent execution models~\cite{alpaca,chain}, allowing safe, synchronous I/O and unsafe, asynchronous I/O.

### 2017-08-09 paper/evaluation.tex
Kiwan's compiler text revised (added questions for Kiwan to respond)

The applications that were used in the evaluation are: (i) \textbf{Bitcount}: counts bits in x y byte pseudo-random string using x algorithms. Each result is compared for correctness; (ii) \textbf{Cold chain}: emulates embedded temperature sensing engine: measurements are randomly generated and are compressed using LZW algorithm for x-entry dictionary and y-byte block; (iii) \textbf{Cuckoo}: implements cuckoo filter---x entry filter filled to y of capacity---program first stores a sequence of pseudo-random numbers, then queries the filter to recover it; (iv) \textbf{RSA}: encrypts predefined x byte long string with a y bytes long key\todo{were keys of 64 bytes tested? If yes: d they exhaust energy}{Amjad}. This set of benchmarks is the same as used in~\cite{chain,alpaca}\todo{Fill-in the missing data and verify its correctness}{Przemek/Amjad}. In addition to Chain benchmarks new applications were implemented: (v) \emph{DFT}, (vi)\emph{Huffman compressor}, and (vii) \emph{Selection Sort}. All applications are compared in Table~\ref{table:benchmark_list} in terms of source lines of code against Chain. \todo{finalize describing all applications once all experiments are done}{Przemek}

@@

\begin{table}[t]
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
		\hline
		Bitcount & Cold chain & Cuckoo & RSA & AR & Sort \\
		\hline\hline
		3 & 2 & 6 & 86 & 34 & > 1 \\
		\hline
	\end{tabular}
	\caption{\sys compile time per application (seconds).}
\label{table:compile_time}
\end{table}

>>>>

The applications that were used in the evaluation are: (i) \textbf{Bitcount}: counts bits in x y byte pseudo-random string using x algorithms. Each result is compared for correctness; (ii) \textbf{Cold chain}: emulates embedded temperature sensing engine: measurements are randomly generated and are compressed using LZW algorithm for x-entry dictionary and y-byte block; (iii) \textbf{Cuckoo}: implements cuckoo filter---x entry filter filled to y of capacity---program first stores a sequence of pseudo-random numbers, then queries the filter to recover it; (iv) \textbf{RSA}: encrypts predefined x byte long string with a y bytes long key\todo{were keys of 64 bytes tested? If yes: d they exhaust energy}{Amjad}. This set of benchmarks is the same as used in~\cite{chain,alpaca}\todo{Fill-in the missing data and verify its correctness}{Przemek/Amjad}. In addition to Chain benchmarks new applications were implemented: (v) \textbf{DFT}, (vi) \textbf{Huffman compressor}, and (vii) \textbf{Selection Sort}. All applications are compared in Table~\ref{table:benchmark_list} in terms of source lines of code against Chain. \todo{finalize describing all applications once all experiments are done}{Przemek}

@@

\begin{table}[t]
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
		\hline
		Bitcount & Cold chain & Cuckoo & RSA & AR & Sort \\
		\hline\hline
		3 & 2 & 6 & 86 & 34 & > 1 \\
		\hline
	\end{tabular}
	\caption{\sys compile time per application (seconds).\todo{for Sort:">" or "<"?}{Kiwan}}
\label{table:compile_time}
\end{table}
